---
title: "Cholera Model"
format: pdf
editor: visual
---

### **1. Setup the SIR Model with Interventions**

We’ll use the `epidemics` package for flexible compartmental modeling. Install it first

```{r}
# Install Epiverse packages
install.packages("epidemics", repos = c("https://epiverse-trace.r-universe.dev", "https://cloud.r-project.org"))
```

### **2. Model Equations with Interventions**

We’ll extend the classic SIR model to include:

-   **Vaccination**: Reduces susceptible population.

-   **Handwashing**: Reduces transmission rate (β).

```{r,echo=FALSE,warning=FALSE}
library(odin)
library(ggplot2)

# Define the model with death rate, vaccination, and handwashing
model <- odin::odin({
  deriv(S) <- -beta * S * I / N - vax_rate * S
  deriv(I) <- beta * S * I / N - gamma * I - mu * I
  deriv(R) <- gamma * I
  deriv(D) <- mu * I  # Deaths from infected compartment
  deriv(V) <- vax_rate * S  # Vaccinated compartment
  
  # Initial conditions
  initial(S) <- S0
  initial(I) <- I0
  initial(R) <- R0
  initial(D) <- 0
  initial(V) <- 0
  
  # Parameters
  beta <- beta0 * (1 - hw_effect)  # Handwashing reduces transmission
  N <- S0 + I0 + R0
  S0 <- user(0)    # Susceptible population (Central Region)
  I0 <- user(0)     # Initial confirmed cases 
  R0 <- user(0)
  gamma <- user(0) # Recovery rate (1/5 days⁻¹)
  mu <- user(0)   # Death rate (19 deaths / 3000 suspected cases ≈ 0.6%)
  beta0 <- user(0) # Baseline transmission rate (r0=2.6)
  hw_effect <- user(0)  # Handwashing effect (baseline = 0)
  
  # Time-varying vaccination rate (starts at day 30)
  vax_rate <- interpolate(
    t_vax, vax_schedule, "constant"
  )
  t_vax[] <- user()  # Time points for vaccination schedule
  vax_schedule[] <- user()  # Vaccination rates at corresponding times
  dim(t_vax) <- user()
  dim(vax_schedule) <- user()
})


```

### **3. Simulate Scenarios**

Compare **no intervention**, **vaccination only**, and **vaccination + handwashing**:

```{r}
# Parameters (based on Central Region data)
parameters <- list(
  S0 = 1e6,       # Estimated Susceptible population
  I0 = 69,        # Initial confirmed cases per report (as of Feb 8, 2025)
  R0 = 0,         # Initial Recovered
  gamma = 0.2,    # Recovery rate
  mu = 0.006,     # Death rate 
  beta0 = 0.52,  # Baseline transmission rate
  hw_effect = 0,  # Handwashing effect (baseline = 0)
  t_vax = c(0, 20, 90),  # Time points for vaccination schedule (day 30 to day 90)
  vax_schedule = c(0,0,0)  # Vaccination rates at time points
)

# Create models with interventions using do.call()
sim_baseline <- do.call(model$new, parameters)

# Vaccination scenario:potency rate of vaccine (50% potency at initial vacination and reduction in potency 40% at day 90) 
sim_vax <- do.call(model$new, modifyList(parameters, list(
  vax_schedule = c(0, 0.50,0.40),# Vaccination starts at day 30
  hw_effect = 0  
)))

# Vaccination + Handwashing (15% transmission reduction)
sim_vax_hw <- do.call(model$new, modifyList(parameters, list(
  vax_schedule = c(0, 0.50, 0.40),  # Vaccination starts at day 30
  hw_effect = 0.1  # 10% transmission reduction
)))

# Run simulations for 120 days
times <- seq(0, 120, by = 1)
out_baseline <- sim_baseline$run(times)
out_vax <- sim_vax$run(times)
out_vax_hw <- sim_vax_hw$run(times)
```

### **4. Visualize Results**

```{r}
# Convert baseline output to data frame
out_baseline_results <- data.frame(
  Time = times,
  Susceptible = out_baseline[, "S"],
  Infected = out_baseline[, "I"],
  Recovered = out_baseline[, "R"],
  Deaths = out_baseline[, "D"],
  Vaccinated = out_baseline[, "V"]
)

# Combine results into a data frame
compare_results <- rbind(
  data.frame(Time = times, Model = "Baseline", Infected = out_baseline[, "I"]),
  data.frame(Time = times, Model = "Vaccination", Infected = out_vax[, "I"]),
  data.frame(Time = times, Model = "Vaccination + Handwashing", Infected = out_vax_hw[, "I"])
)

# Plot
# Plot results
ggplot(out_baseline_results, aes(x = Time)) +
  geom_line(aes(y = Susceptible, color = "Susceptible"), linewidth = 1.2) +
  geom_line(aes(y = Infected, color = "Infected"), linewidth = 1.2) +
  geom_line(aes(y = Recovered, color = "Recovered"), linewidth = 1.2) +
  labs(
    title = "Cholera Outbreak Stimulation in Central Region, Ghana",
    subtitle = "SIR Model with Deaths, Vaccination (starts at day 30), and Handwashing (10% reduction)",
    y = "Number of Individuals",
    x = "Time (Days)",
    color = "Compartment"
  ) +
  theme_minimal()

ggplot(compare_results, aes(x = Time, y = Infected, color = Model)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Cholera Outbreak Stimulation in Ghana: Impact of Interventions",
    subtitle = "SIR Model with Vaccination and Handwashing Interventions",
    y = "Active Infections",
    caption = "Assumptions: R0 = 2.5, 5-day infectious period, 120-day simulation"
  ) +
  theme_minimal()
```
